\subsection{Soundness and Completeness Proof}\label{soundness}

Given an arbitrary list of constraints, the unification algorithm
either fails or returns its most general unifier. We have the
following properties:
\begin{itemize}
    \item Soundness: the substitution produced is a unifier of the constraints.

    \item Completeness: the returned substitution is the least
      unifier, according to the substitution ordering defined in
      Section \ref{substitution}.
\end{itemize}

A substitution $S$ is called a unifier of a list of constraints
$\mathbb{C}$ according to whether $\unifier(\mathbb{C},S)$ is provable,
where $\unifier(\mathbb{C},S)$ is defined by induction on $\mathbb{C}$
as follows:

\[ \begin{array}{lcl}
     \unifier([],S)                             & = & \True \\
     \unifier((\tau\eeq\tau') :: \mathbb{C}',S) & = & S(\tau) = S(\tau') \land \unifier(\mathbb{C}',S)
  \end{array} 
\]

\newcommand{\least}{\text{\it least\/}}

A substitution $S$ is a most general unifier of a list of constraints
$\mathbb{C}$ if, for any other unifier $S'$ of $\mathbb{C}$, there
exists $S_1$ such that $S' = S_1 \circ S$; formally:

\[
  \least(S,\mathbb{C}) = 
    \forall S'.\,\unifier(\mathbb{C},S') \to 
       \exists S_1.\, \forall \alpha.\,(S_1 \circ S)(\alpha) = S'(\alpha)
\]
The type of the unification function is a dependent type that ensures
the following property of the returned substitution $S$:

\[ \bigl(\unifier(\mathbb{C},S) \land 
    \least(S,\mathbb{C})\bigr) \lor \text{UnifyFailure($\mathbb{C}$)}
\]
where UnifyFailure($\mathbb{C}$) is a type that encodes the reason why
unification of $\mathbb{C}$ fails. There are two possible causes of
failure: 1) an occurs check error, 2) an error caused by trying to
unify distinct type constructors.

In the formalization source code, the definition of the $\unify$ function
contains ``holes''\footnote{A
  hole in a function definition is a subterm that is left
  unspecified. In Coq, holes are represented by underscores and such
  unspecified parts of a definition are usually filled by tactic
  generated terms.} to mark positions where proof terms are
expected. Instead of writing such proof terms, we left them
unspecified and use tactics to fill them with appropriate proofs. In
the companion source code, the unification function is full of such
holes and they mark the position of proof obligations for soundness,
completeness and termination for each equation of the definition of
$\unify$.

In order to prove soundness obligations we define several small lemmas
that are direct consequences of the definition of the application of
substitutions, which are omitted for brevity. Other lemmas necessary
to ensure soundness are sketched below. They specify properties of
unification and application of substitutions.

\begin{Lemma}\label{unifyty}
For all type variables $\alpha$, types $\tau,\tau'$ and substitutions
$S$, if $S(\alpha) = S(\tau')$ then $S(\tau) = S([\alpha \mapsto
  \tau']\,\tau)$.
\end{Lemma}
\begin{proof}
Induction over the structure of $\tau$.
\end{proof}

\begin{Lemma}
For all type variables $\alpha$, types $\tau$, variable contexts
$\mathcal{V}$ and constraint sets $\mathbb{C}$, if $S(\alpha) = S(\tau)$
and $\unifier(\mathbb{C},S)$ then
$\unifier([\alpha\mapsto\tau]\,\mathbb{C},S)$.
\end{Lemma}
\begin{proof}
Induction over $\mathbb{C}$ using Lemma \ref{unifyty}.
\end{proof}

Completeness proof obligations are filled by scripted automatic proof tactics
using Lemma \ref{extsubst}.

%\input{tactics}
